Whole Genome Resequencing for Population Genomics

### 2022-08-17 RADseq.Rmd
Fixed references in RADseq, updated fastq2vcf to most recent version from SW

"Restriction-site   Associated   DNA   sequencing"  - RADseq - combines  restriction  enzymes  with next-gen, massively parallel,  short-read  sequencing. RADseq involves the use of restriction enzymes, which are used to shear DNA at restriction enzyme cutsites. RADseq comes in different flavors. Double-digest RADseq (ddRAD; [@Peterson2012-tr])  selects  markers  with  two  restriction enzymes  with  different  cut  frequencies. Genotype-by-Sequencing  (GBS;  [@Elshire2011-or])  uses  a  frequent-cutting  restriction  enzyme  with PCR size selection. There are a number of reviews comparing different RADseq and GBS methods (e.g., [@Andrews2016-ay]). In this guide we use the term "RADseq" to refer to any of these protocols, including those that don't involve random shearing of data. When recommendations are specific to a certain type of RAD/GBS, we will explicitly say so.

@@

|  | [dDocent](https://www.ddocent.com/) | [ipyrad](https://ipyrad.readthedocs.io/en/latest/index.html)| [Stacks2](https://catchenlab.life.illinois.edu/stacks/) |
|---|---|---|---|
| Supported datatypes | Paired-end: ddRAD, ezRAD, RAD (random shearing), data with large overlap between forward and reverse reads; Single-end: any RADseq method  de novo and reference-based. If doing de novo assembly, reads cannot be trimmed outside of dDocent | Wide range of RAD/GBS methods with paired/single end data, see [ipyrad documentation](https://ipyrad.readthedocs.io/en/master/4-data.html) | Paired-end and single-end data for single and double digest RAD and DART; GBS with single-end sequencing. NOT suitable for paired-end GBS. |
| Unique aspects | Novel data reduction approach used to inform coverage cutoffs | Python API for popgene and phylogenetic analyses written specifically for RAD data |  |
| Documentation | Very good (esp. tutorials), active community support on Google Groups | Excellent (esp. installation, parameter explanations, and tutorials), active community support on Gitter | Very good, active community support on Google Groups |
| Speed/Accuracy | fastest and most accurate | close to dDocent accuracy, more over-splitting is possible  if parameters are not tuned | produces some untrue genome fragments (esp. with higher levels of indel polymorphism), but can be addressed with downstream filtering |
| Open source/development | Open source | Open source, primarily combines existing software | No |
| Filtering options | Minimal default filtering as implemented in VCFtools. Ideal for those who want full freedom in filtering their SNPs. | Wide range of filtering options  | Moderate range of filtering options |
| Output options | Only produces a VCF, fasta of the de novo assembly, and individual BAM files  from mapping to the reference  | Lots of output formats (see [documentation](https://ipyrad.readthedocs.io/en/master/output_formats.html)) | Lots of Stacks-specific output and log files, as well as various inputs for popgen programs (see [documentation](https://catchenlab.life.illinois.edu/stacks/comp/populations.php)) |
| Popgen analyses | No additional popgen analyses included  | Unique, flexible Python API implementing range of popgen and phylogenetic methods (see [documentation](https://ipyrad.readthedocs.io/en/master/API-analysis/index.html))   | Popgen summary statistics (F-statistics, pi), including kernel-smoothing along reference genome. Underlying equations and assumptions for these are not very clear. |
|  |  |  |  |

Some groups have also developed pipelines for specific flavors of RAD (e.g., [Matz lab for 2b-RAD](https://github.com/z0on/2bRAD_denovo/blob/master/2bRAD_README.sh)) or proprietary software (eg, [DARTseq](http://georges.biomatix.org/storage/app/media/uploaded-files/dartR_Workbook.pdf)), which we touch on below. **In most cases, it is recommended that you use a pipeline developed and tested for RAD data, especially if you are making a de novo assembly.** [@LaCava2020-bg] have an excellent study where they review various *de novo* assemblers that are used in these pipelines. Assemblers that were not explicitly developed for short reads (eg, Velvet, ABySS) performed very poorly, while CD-HIT (the assembler in dDocent) performed the best. Stacks/Stacks2 and vsearch (the assembler in ipyrad) both performed worse when analyzing simulated data with indels. It should be noted that all of these RAD pipelines have optimized the specific parameters of these de novo assemblers to work with RAD data of various flavors, which may not be fully reflected in the LaCava study. Based on the MarineOmics RAD panel, the general consensus is that all the popular pipelines can produce adequately accurate datasets (with appropriate parameter optimization and data filtering).     
![Table 3 from LaCava et al 2019. CD-HIT is used in dDocent, VSEARCH is used in ipyrad.](images/lacavaTable3.png)

@@

It looks like samples X,Y,Z have a lot of missing data relative to other samples. We can remove them from our dataset for now, and just keep exploring our data, but if you decide to exclude them permanently from the analysis you should eventually rerun your genotyping pipeline without those samples, esp. if doing a de novo assembly.

2) Another great way to explore missingness with your data is to construct a heatmap of loci that are genoptyped between pairs of samples. Generally, samples that are more closely related will share more loci with each other due to cutsite dropout (cite). Deviations from that pattern can pinpoint bad samples as well as batch effects.
  + [ipyrad API tutorial](https://ipyrad.readthedocs.io/en/latest/API-analysis/cookbook-sharing.html) for making a locus sharing heatmap, requires ipyrad output or ipyrad's vcf2hdf5 conversion  
  + [code to generate locus sharing/missingness heatmap from a VCF](https://github.com/atcg/clustOpt)

@@

One of the most powerful methods for exploring your data is a Principle Components Analysis. From [this genetics tutorial](https://comppopgenworkshop2019.readthedocs.io/en/latest/contents/03_pca/pca.html): "To understand how PCA works, consider a single individual and its representation by its 593,124 markers. Formally, each individual is a point in a 593,124-dimensional space, where each dimension can take only the three possible genotypes indicated above, or have missing data. To visualize this high-dimensional dataset, we would like to project it down to two dimensions. But as there are many ways to project the shadow of a three-dimensional object on a two dimensional plane, there are many (and even more) ways to project a 593,124-dimensional cloud of points to two dimensions. What PCA does is figuring out the “best” way to do this project in order to visualise the major components of variance in the data." See [here](https://michaelaalcorn.medium.com/yet-another-pca-explanation-727dff6ce26a) for a linear algebra-based explanation of PCA).

@@

Here we use SeqArray and SNPRelate to run a PCA in R.

@@

*Reminder:* Missing data is a feature of RAD. Be aware of how different analysis tools deal with missingness, especially PCA which will fill in all missing data with some values. Here, the PCA from SNPRelate imputes missing genotypes as the dosage mean.

@@

* Do any observed clusters match a source of batch effect? [how to remove batch effects](#batch)

@@

Nicolas Lou and Nina Therkildsen have a [great preprint](https://www.authorea.com/users/380682/articles/532568-batch-effects-in-population-genomic-studies-with-low-coverage-whole-genome-sequencing-data-causes-detection-and-mitigation) and [Github repo](https://github.com/therkildsen-lab/batch-effect) discussing causes, detection, and mitigation of batch effects in low coverage WGS, where batch effects can have particularly large effects. While some of their recommendations are specific to reference-based analyses, others are relevant to RAD methods too (eg, batch effects due to different sequencing platforms, DNA degradation levels, and sequencing depth).

@@

You can also check heterozygosity. Large differences in heterozygosity between outliers can suggest different species or cryptic hybrids.

<!--Note: Two outliers are obvious.
Large differences in heterozygosity between samples are clear, with one location showing two distinct clusters of samples. These samples were later determined using mtDNA and morphological analysis to be cryptic species.
*Unpublished data from a marine species in Australian waters.*-->

You can then estimate genetic divergence between these clusters using pairwise Fst and/or phylogenetic inference to determine how diverged the groups are. If they are very diverged, it may be better to assemble each cryptic species separately, especially for denovo assemblies with dDocent and Stacks, especially since often times restriction cut-sites are not always conserved; see [@Wagner2013-ff,@Rubin2012-hw]. These two pipelines used a set of given samples to create a "reference" or "catalog" to which all other samples are mapped. If highly diverged samples are included in creating this reference, it can lead to [oversplitting](#cluster).

@@

Mapping defaults for popular RADseq pipelines:
L: clipping penalty for 5' and 3'
A: matching score
O gap open penalty
T: mapping quality filter

dDocent: bwa mem -L 20,5 -t 8 -a -M -T 10 -A1 -B 3 -O 5 -R

* can change -A and -O during the dDocent pipeline.    
ipyrad: (defaults) -L 5,5 -B 4 -A1 -O 6 -T 30

@@

Our whole genome resequencing section has some good content on mapping(link). General considerations with RAD data:

@@

##### 2) Require higher read depth thresholds for assembly and genotyping.  
For RADseq, ideally you are generating an average of *at least* 10-20 sequence reads (10x-20x coverage) for each RAD locus in your library preparation (but this doesn't always happen...). For diploid species, this would be sequencing each RAD haplotype at least 5x-10x. The SNP calling step of RAD pipelines all require setting a minimum number of reads to determine whether a site is homozygous or heterozygous. If this threshold is too low, sequencing errors are more likely to be included and the SNP caller will have a harder time accurately identifying SNPs ([in which case, you'd be better off with genotype likelihood methods](https://github.com/therkildsen-lab/genomic-data-analysis/blob/master/lcwgs_data_analysis.md). If it's too high, then real polymorphic loci will be excluded. All of the three major pipelines allow users to implement read depth thresholds

Generating multiple sequencing reads per locus is helpful when calling SNPs in order to  distinguishing between reads with sequencing errors and.

@@

`r colorize("This section is still under development! Please contribute on Github or provide feedback in [Discussions](https://github.com/MarineOmics/marineomics.github.io/discussions)", "red")`

@@

* Don't over filter for missingness
 + papers showing how too much filtering can skew results  
  + Empirical: Diaz-Arce 2019, Tripp 2017, Silliman 2021, Eaton 2017, Martin-Hernanz (2019)
  + Simulations: Huang & Knowles 2016, Leaché et al., 2015, Gautier et al., 2012, Rivera-Colón et al., 2021

* some ways to account for missing data in downstream analyses
 + impute (PCA), subsample (PCA, structure, pretty much anything with SNPs), try a few different missing filters and see what impact it has on results

@@

## Repeat as needed {#repeat}
# Helpful tutorials  
* [Marine Genomics open course](https://baylab.github.io/MarineGenomics/)
* [UNIX tutorial](http://evomics.org/learning/unix-tutorial/)
* [what is a VCF file and how can I filter it?](http://evomics.org/learning/population-and-speciation-genomics/2020-population-and-speciation-genomics/first-steps-in-genomic-data-analysis/)

>>>>

"Restriction-site   Associated   DNA   sequencing"  - RADseq - combines  restriction  enzymes  with next-gen, massively parallel,  short-read  sequencing. RADseq involves the use of restriction enzymes, which are used to shear DNA at restriction enzyme cutsites. RADseq comes in different flavors. Double-digest RADseq (ddRAD; @Peterson2012-tr)  selects  markers  with  two  restriction enzymes  with  different  cut  frequencies. Genotype-by-Sequencing  (GBS;  @Elshire2011-or)  uses  a  frequent-cutting  restriction  enzyme  with PCR size selection. There are a number of reviews comparing different RADseq and GBS methods (e.g., @Andrews2016-ay). In this guide we use the term "RADseq" to refer to any of these protocols, including those that don't involve random shearing of data. When recommendations are specific to a certain type of RAD/GBS, we will explicitly say so.

@@

|  | [dDocent](https://www.ddocent.com/) | [ipyrad](https://ipyrad.readthedocs.io/en/latest/index.html)| [Stacks2](https://catchenlab.life.illinois.edu/stacks/) |
|---|---|---|---|
| Supported datatypes | Paired-end: ddRAD, ezRAD, RAD (random shearing), data with large overlap between forward and reverse reads; Single-end: any RADseq method  de novo and reference-based. If doing de novo assembly, reads cannot be trimmed outside of dDocent | Wide range of RAD/GBS methods with paired/single end data, see [ipyrad documentation](https://ipyrad.readthedocs.io/en/master/4-data.html) | Paired-end and single-end data for single and double digest RAD and DART; GBS with single-end sequencing. NOT suitable for paired-end GBS. |
| Unique aspects | Novel data reduction approach used to inform coverage cutoffs | Python API for popgen and phylogenetic analyses written specifically for RAD data | `populations` module calculates sliding window and site-specific popgen metrics |
| Documentation | Very good (esp. tutorials), active community support on Google Groups | Excellent (esp. installation, parameter explanations, and tutorials), active community support on Gitter | Very good, in depth tutorials as published manuscripts, active community support on Google Groups |
| Speed/Accuracy (based on discussion in [panel seminar](https://www.youtube.com/watch?v=C74GBESeIq4)) | fastest and most accurate | close to dDocent accuracy, more over-splitting is possible  if parameters are not tuned | produces some untrue genome fragments (esp. with higher levels of indel polymorphism), but can be addressed with downstream filtering |
| Open source/development | Open source, primarily combines existing software | Open source | No |
| Filtering options | Minimal default filtering as implemented in VCFtools. Ideal for those who want full freedom in filtering their SNPs. | Wide range of filtering options, some hard-coded filtering to deal with paralogs  | Moderate range of filtering options |
| Output options | Only produces a VCF, fasta of the de novo assembly, and individual BAM files  from mapping to the reference  | Lots of output formats (see [documentation](https://ipyrad.readthedocs.io/en/master/output_formats.html)) | Lots of Stacks-specific output and log files, as well as various inputs for popgen programs (see [documentation](https://catchenlab.life.illinois.edu/stacks/comp/populations.php)) |
| Popgen analyses | No additional popgen analyses included  | Unique, flexible Python API implementing range of popgen and phylogenetic methods (see [documentation](https://ipyrad.readthedocs.io/en/master/API-analysis/index.html))   | Popgen summary statistics (F-statistics, pi), including kernel-smoothing along reference genome. Underlying equations and assumptions for these are not very clear. |
|  |  |  |  |

Some groups have also developed pipelines for specific flavors of RAD (e.g., [Matz lab for 2b-RAD](https://github.com/z0on/2bRAD_denovo/blob/master/2bRAD_README.sh)) or proprietary software (eg, [DARTseq](http://georges.biomatix.org/storage/app/media/uploaded-files/dartR_Workbook.pdf)). **In most cases, it is recommended that you use a pipeline developed and tested for RAD data, especially if you are making a de novo assembly.** [@LaCava2020-bg] have an excellent study where they review various *de novo* assemblers that are used in these pipelines. Assemblers that were not explicitly developed for short reads (eg, Velvet, ABySS) performed very poorly, while CD-HIT (the assembler in dDocent) performed the best. Stacks/Stacks2 and vsearch (the assembler in ipyrad) both performed worse when analyzing simulated data with indels. It should be noted that all of these RAD pipelines have optimized the specific parameters of these de novo assemblers to work with RAD data of various flavors, which may not be fully reflected in the LaCava study.

Based on the MarineOmics RAD panel, the general consensus is that all the popular pipelines can produce adequately accurate datasets (with appropriate parameter optimization and data filtering).     
![Table 3 from LaCava et al 2019. CD-HIT is used in dDocent, VSEARCH is used in ipyrad.](images/lacavaTable3.png)

@@

It looks like samples OR3_13_C3,CA4_12_C1, OR3_3_C7, CA4_13_C1, BC2_6_C2, BC2_9_C4 have a lot of missing data relative to other samples. We can remove them from our dataset for now, and just keep exploring our data, but if you decide to exclude them permanently from the analysis you should eventually rerun your genotyping pipeline without those samples, esp. if doing a de novo assembly.

2) Another great way to explore missingness with your data is to construct a heatmap of loci that are genoptyped between pairs of samples. Generally, samples that are more closely related will share more loci with each other due to cutsite dropout. Deviations from that pattern can pinpoint bad samples as well as batch effects.
  + [ipyrad API tutorial](https://ipyrad.readthedocs.io/en/latest/API-analysis/cookbook-sharing.html) for making a locus sharing heatmap, requires ipyrad output or ipyrad's vcf2hdf5 conversion  
  + [code to generate locus sharing/missingness heatmap from a VCF](https://github.com/atcg/clustOpt)

@@

One of the most powerful methods for exploring your data is a Principle Components Analysis. From [this genetics tutorial](https://comppopgenworkshop2019.readthedocs.io/en/latest/contents/03_pca/pca.html): "To understand how PCA works, consider a single individual and its representation by its 593,124 markers. Formally, each individual is a point in a 593,124-dimensional space, where each dimension can take only the three possible genotypes indicated above, or have missing data. To visualize this high-dimensional dataset, we would like to project it down to two dimensions. But as there are many ways to project the shadow of a three-dimensional object on a two dimensional plane, there are many (and even more) ways to project a 593,124-dimensional cloud of points to two dimensions. What PCA does is figuring out the “best” way to do this project in order to visualise the major components of variance in the data." See [here](https://michaelaalcorn.medium.com/yet-another-pca-explanation-727dff6ce26a) for a linear algebra-based explanation of PCA.

@@

Here we use _SeqArray_ and _SNPRelate_ to run a PCA in R.

@@

*Reminder:* Missing data is a feature of RAD. Be aware of how different analysis tools deal with missingness, especially PCA which will fill in all missing data with some values. Here, the PCA from _SNPRelate_ imputes missing genotypes as the mean genotype across samples.

@@

* Do any observed clusters match a source of batch effect? [How to remove batch effects](#batch)

@@

Nicolas Lou and Nina Therkildsen have a [great article](https://onlinelibrary.wiley.com/doi/abs/10.1111/1755-0998.13559) and [Github repo](https://github.com/therkildsen-lab/batch-effect) discussing causes, detection, and mitigation of batch effects in low coverage WGS, where batch effects can have particularly large effects [@@Lou2021-ew]. While some of their recommendations are specific to reference-based analyses, others are relevant to RAD methods too (eg, batch effects due to different sequencing platforms, DNA degradation levels, and sequencing depth).

@@

You can also check heterozygosity. Large differences in heterozygosity between outliers can suggest different species or cryptic hybrids. Two outliers are obvious in the previous PCA. Large differences in heterozygosity between samples are clear, with one location showing two distinct clusters of samples. These samples were later determined using mtDNA and morphological analysis to be cryptic species.
*Unpublished data from a marine species in Australian waters.*

<!--Note: -->

You can then estimate genetic divergence between these clusters using pairwise Fst and/or phylogenetic inference to determine how diverged the groups are. If they are very diverged, it may be better to assemble each cryptic species separately, especially for denovo assemblies with dDocent and Stacks, especially since often times restriction cut-sites are not always conserved; see [@Wagner2013-ff; @Rubin2012-hw]. These two pipelines used a set of given samples to create a "reference" or "catalog" to which all other samples are mapped. If highly diverged samples are included in creating this reference, it can lead to [oversplitting](#cluster).

@@

Mapping defaults for popular RADseq pipelines:  
`L`: clipping penalty for 5' and 3'  
`A`: matching score  
`O`: gap open penalty  
`T`: mapping quality filter

dDocent: `bwa mem -L 20,5 -t 8 -a -M -T 10 -A1 -B 3 -O 5 -R`

* can change -A and -O interactively during the dDocent pipeline.

ipyrad: (defaults) `-L 5,5 -B 4 -A1 -O 6 -T 30`

@@

Our [whole genome resequencing tutorials](https://marineomics.github.io/WGS_intro.html)] has some good content on mapping. General considerations with RAD data:

@@

##### 2) Require higher read depth thresholds for assembly and genotyping.  
For RADseq, ideally you are generating an average of *at least* 10-20 sequence reads (10x-20x coverage) for each RAD locus in your library preparation (but this doesn't always happen...). For diploid species, this would be sequencing each RAD haplotype at least 5x-10x. The SNP calling step of RAD pipelines all require setting a minimum number of reads to determine whether a site is homozygous or heterozygous. If this threshold is too low, sequencing errors are more likely to be included and the SNP caller will have a harder time accurately identifying SNPs ([in which case, you'd be better off with genotype likelihood methods](https://github.com/therkildsen-lab/genomic-data-analysis/blob/master/lcwgs_data_analysis.md). If it's too high, then real polymorphic loci will be excluded. The *too high* threshold for your dataset can usually best be determined by using a few different read depth thresholds and comparing how the results change downstream analyses.

@@

`r colorize("Tutorials for this section are still under development! Please contribute on Github or provide feedback in [Discussions](https://github.com/MarineOmics/marineomics.github.io/discussions)", "red")`

@@

* Don't over filter for missingness
    + Papers showing how too much filtering can skew results:
       + Empirical:  [@Eaton2017-gx; @Tripp2017-cp; @Martin-Hernanz2019-bu; @Silliman2021-oo; @Diaz-Arce2019-nh]  
       + Simulation investigating effects and sources of missing data: [@Huang2016-na; @Leache2015-eg; @Gautier2013-vj; @Rivera-Colon2021-en]
* Some ways to account for missing data in downstream analyses:
   + impute mising data (PCA), subsample (PCA, structure, pretty much anything with SNPs), try a few different missing filters and see what impact it has on results

Jon Puritz has a [great tutorial](http://www.ddocent.com/filtering/) on filtering SNPs from RADseq data assembled with dDocent. While it requires certain VCF fields to be present that aren't generated by ipyrad or Stacks, it contains great concepts on how to filter.

@@

## Repeat as needed {#repeat}

# Helpful tutorials  
* [Marine Genomics open course](https://baylab.github.io/MarineGenomics/)
* [UNIX tutorial](http://evomics.org/learning/unix-tutorial/)
* [what is a VCF file and how can I filter it?](http://evomics.org/learning/population-and-speciation-genomics/2020-population-and-speciation-genomics/first-steps-in-genomic-data-analysis/)

### 2022-08-17 WGS_intro.Rmd
Added a ref
Fixed references in RADseq, updated fastq2vcf to most recent version from SW

* [Fastq-to-VCF SnakeMake pipeline](fastq2vcf_pipeline.html): in-depth explanation of an automated short-read mapping and variant calling pipeline maintained by [Harvard Informatics](https://github.com/harvardinformatics/shortRead_mapping_variantCalling). Useful if you plan to adopt this pre-packaged automated and parallelizable pipeline, would like to understand its different components but not necessarily change it substantially.

@@

* [Detection and mitigation of batch effects](https://github.com/therkildsen-lab/batch-effect/blob/main/tutorial/tutorial.md): a tutorial for the detection and mitigation of batch effects with low-coverage WGS data, with example datasets and hands-on exercises. It is associated with the paper [(Lou and Therkildsen 2021)](https://onlinelibrary.wiley.com/doi/abs/10.1111/1755-0998.13559).

>>>>

* [Fastq-to-VCF SnakeMake pipeline](fastq2vcf_june2022_update.html): in-depth explanation of an automated short-read mapping and variant calling pipeline maintained by [Harvard Informatics](https://github.com/harvardinformatics/shortRead_mapping_variantCalling). Useful if you plan to adopt this pre-packaged automated and parallelizable pipeline, and would like to understand its different components, but not necessarily change it substantially.

@@

* [Detection and mitigation of batch effects](https://github.com/therkildsen-lab/batch-effect/blob/main/tutorial/tutorial.md): a tutorial for the detection and mitigation of batch effects with low-coverage WGS data, with example datasets and hands-on exercises. It is associated with the paper [@Lou2021-ew](https://onlinelibrary.wiley.com/doi/abs/10.1111/1755-0998.13559).
